---
title: "scDist: Robust identification of perturbed cell types in single-cell RNA-seq data"
author: "R package version 1.0.0"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## System Requirements

`R` is required to use `scDist`. In development, `R` version 4.0.0 and greater were used, but there may be compatibility with previous versions. 

## Installation 

From the R console, `devtools::install_github("phillipnicol/scDist")`. Installation should take less than a minute on a standard machine. 

## Demo 

The input to `scDist` is a normalized count matrix and correpsonding metadata that describes what condition and patient each cell belongs to. In this demo, we create a simulated dataset with 10 cell types. The demo should take less than a minute to run on a standard machine. The code is also modifiable to see how `scDist` performs for different parameter values.

```{r}
library(scDist)
set.seed(1126490984)
```

Generate simulated normalized counts with a distance (separation) $D$ and patient-specific effect $\tau$: 

```{r}
simCellType <- function(D,tau,nn=100,G=1000,N1=5,N2=5,J=30,label="A") {
  beta_true <- rep(0,G)
  beta_true[sample(1:G,size=nn)] <- uniformly::runif_on_sphere(n=1,d=nn,r=D)[1,]
  y <- matrix(0, nrow=(N1+N2)*J,ncol=G)
  for(i in 1:G) {
    cntr <- 1
    for(j in 1:N1) {
      omega <- rnorm(1,mean=0,sd=tau)
      y[cntr:(cntr+J-1),i] <- omega+rnorm(J)
      cntr <- cntr+J
    }
    for(j in 1:N2) {
      omega <- rnorm(1,mean=0, sd=tau)
      y[cntr:(cntr+J-1),i] <- beta_true[i]+omega+rnorm(J)
      cntr <- cntr+J
    }
  }
  
  response <- rep(0,(N1+N2)*J)
  response[1:(N1*J)] <- 1

  samples <- c()
  for(i in 1:(N1+N2)) {
    samples <- c(samples, rep(i,J))
  }

  meta.data <- data.frame(response=response,patient=as.factor(samples),clusters=label)
  out <- list()
  out$Y <- t(y); out$meta.data <- meta.data
  return(out)
}

simData <- function(nct=10, J=50, N1, N2, G=1000, nn=100,tau=0.5) {
  Y <- matrix(0,nrow=G,ncol=0)
  meta.data <- data.frame(response=NULL,
                          patient=NULL,
                          clusters=NULL)
  D.true <- rep(0,nct)
  for(i in 1:nct) {
    D.true[i] <- rexp(n=1,rate=0.05)
    out <- simCellType(D=D.true[i],J=rpois(n=1,lambda=J),N1=N1,N2=N2,label=letters[i],tau=tau)
    Y <- cbind(Y,out$Y)
    meta.data <- rbind(meta.data,out$meta.data)
  }
  
  out$Y <- Y
  out$meta.data <- meta.data
  out$D.true <- D.true
  return(out)
}

data <- simData(N1=5,N2=5)
```

The `meta.data` data frame contains the assignment of each cell to condition, patient, and cell type:

```{r}
head(data$meta.data)
```

Now we apply scDist: 

```{r}
out <- scDist(data$Y,data$meta.data,fixed.effects = "response",
              random.effects="patient",
              clusters="clusters")
```

The results data frame gives a summary of the estimated distance and uncertainty for each cell type 

```{r}
out$results
```

The true distances are

```{r}
names(data$D.true) <- letters[1:length(data$D.true)]
data$D.true
```


We can also plot the results

```{r}
DistPlot(out)
```

## Reference

If you use `scDist` in your work, please cite:

Nicol, P.B., Paulson, D., Qian, G., Liu, X.S., Irizarry, R.A., and Sahu, A.D. (2023). Robust identification of perturbed cell types in single-cell RNA-seq data. bioRxiv.


